Ecological management of stochastic systems with long transients

### 2019-12-30 manuscript/manuscript.Rmd
tidy up and WIP on manuscript

---
title: "Consequences of long transients in model inference and optimal control"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  I explore the impact of long transient dynamics for model inference and optimal
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article

@@

```{r figure1, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter a, the system goes from bistable (red, a = 28.5) to a ghost attractor (blue, a = 27). For both curves, r = 0.8, K = 153, Q = 2, H = 20."}
c(`ghost attractor` = 27,           
  `bistable attractor` = 28.5) %>%
  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
  ggplot(aes(x, f)) + 
    geom_line(aes(lty = scenario, col = scenario), lwd = 1) +
    geom_hline(aes(yintercept = 0)) + 
    labs(y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
    scale_color_solarized()

@@

Under appropriate choice of parameters, the model is bistable, with two stable equilibiria at non-zero population sizes -- a low, "endemic" state, and a much higher "outbreak" state.  In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emergign from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses. This saturating response and the resulting threshhold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g.@Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate. Initially this results in the strengthening (concave up) portion of the S curve as other defences are induced by the attack.  However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defences, leading to a saturating response.  Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.  Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defences of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting avaliable tree habitat $K$ [@Harvey2014].

Thus our simple dynamical model from \eqref{model} captures the qualitative processes and dynamics expected in the beetle outbreak system. Bear in mind that this model is not intended to be a precise, quantitive match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.

We also observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

Figure 2 illustrates several simulations from the model, starting from a low density, under both the scenario of a stable endemic state and a ghost attractor.  Note that it can in general be quite difficult to distinguish dynamics of the ghost from the dynamics of the stable attractor.

@@

```{r}
P_ghost <- transition_matrix(states, actions, may(27.2), p$sigma)
P_weak_attractor <- transition_matrix(states, actions, may(28), p$sigma)
P_bistable <- transition_matrix(states, actions, may(28), p$sigma)
```

@@

```{r}
x0 <- which.min(abs(states - p$x0))
Tmax <- 200
set.seed(12345)
reps <- 50
ghost_sim <- map_dfr(1:reps, function(i)
  sim(P_ghost, x0, Tmax)  %>% mutate(state = states[state], scenario = "ghost"),
  .id = "rep")
bistable_sim <- map_dfr(1:reps, function(i)
  sim(P_weak_attractor, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
  .id = "rep")

bind_rows(ghost_sim, bistable_sim) %>%
  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
```

@@

$$\max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)$$
where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Marescot2013].

@@

## Optimal Planning under transient dynamics vs stable state dynamics

@@

```{r}
soln_ghost <- mdplearning::mdp_compute_policy(list(P_ghost), reward, discount, max_iter = 1e4, epsilon = 1e-2)
soln_bistable <-mdplearning::mdp_compute_policy(list(P_bistable), reward, discount, max_iter = 1e4, epsilon = 1e-4)

#soln_ghost <-MDPtoolbox::mdp_value_iteration(P_ghost, reward, discount, max_iter = 5e3, epsilon = 1e-5)

@@

```{r}
## With more noise, these are more similar.  With smaller noise, the ghost policy remains aggressive over a larger range
tibble(state = states,
       ghost = actions[soln_ghost$policy],
       bistable =  actions[soln_bistable$policy]
       ) %>%
  pivot_longer(c(ghost, bistable), 
               names_to="model",
               values_to = "action") %>%
  ggplot(aes(state,action, color = model)) + 
  geom_point(alpha=0.6) + 
  scale_color_solarized()
```

@@

```{r}
fifty_fifty <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
                                             model_prior = c(0.5, 0.5),
                                             max_iter = 1e4, epsilon = 1e-2)

@@

```{r}

@@

@@

```{r}
 bind_rows(ghost = sim_ghost, 
           bistable = sim_bistable, 
           fifty_fifty = sim_fiftyfifty, 
           seventy_thirty = sim_seventythirty, .id = "prior")  %>% 
  mutate(state = states[state], action = actions[action]) %>% 
  select(-obs,-value) %>%
  pivot_longer(c(-time, -prior, -reps)) %>%
  ggplot(aes(time, value, group = interaction(name, reps), col=name)) + 
  geom_path(alpha=0.5) + facet_wrap(~prior)
```

@@

```{r}
set.seed(12345)
Tmax <- 50
learning <- mdp_learning(list(P_ghost, P_bistable), reward, discount, 
                   model_prior = c(0.5, 0.5),
                   x0 = x0, Tmax = Tmax,
                  true_transition = P_ghost,
                  max_iter = 1e4, epsilon = 1e-2)

```{r}
 learning$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r}
model_names <- c("Ghost", "Bistable")
learning$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% floor(seq(1,10, length.out=5))) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = model_names[as.integer(param)]) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1])
```

```{r}
```

## Long transients when the underlying model is not known

One of the more intuitive implications of the existence of long transients is the inherent risk in mistaking them for stable state dynamics.  For example, Figure \ref{armina} illustrates how a transient

### Stochastic dynamics and transients

Stochastic dynamics can weaken the average influence of a ghost attractor, but long transients are still likely.  Stochastic dynamics can also make weak attractors behave essentially as ghost attractors.

@@

Stochastic dynamic programming solutions were computed in R [@R] using value iteration using the package `MDPToolbox` [@Marescot2013; @MDPtoolbox].  Code to reproduce this analysis is available at <https://github.com/cboettig/decisions-vs-transients>

>>>>

---
title: "Ecological management of stochastic systems with long transients"
date: "`r Sys.Date()`"
author:
  - name: Carl Boettiger
    email: cboettig@berkeley.edu
    affiliation: ucb
#    footnote: Corresponding Author
address:
  - code: ucb
    address: "ESPM Department, University of California, 130 Mulford Hall Berkeley, CA 94720-3114, USA"
abstract: |
  Transient I explore the impact of long transient dynamics for model inference and optimal
journal: Theoretical ecology
layout: 3p
bibliography: refs.bib
output: rticles::elsevier_article

@@

```{r fig_model, fig.cap="Change in population size $X_{t+1} - X_t = f(X_t)$. Equilibria occur when the change curve crosses zero, stable points arise at crossings with negative slopes, unstable equilibria have positive slopes.  By changing parameter $a$, the system goes from bistable (red, $a = 28$) to a ghost attractor (blue, $a = 27.2$). For both curves, $r = 0.8$, $K = 153$, $Q = 2$, $H = 20$."}
c(`ghost attractor` = 27.2,           
  `bistable attractor` = 28) %>%
  map_dfr(function(a) tibble(x = states, f = may(a)(x,0) - x, a = a), .id = "scenario") %>% 
  ggplot(aes(x, f)) + 
    geom_line(aes(lty = scenario, col = scenario), lwd = 1) +
    geom_hline(aes(yintercept = 0)) + 
    labs(y = bquote(X[t+1] - X[t]), x = bquote(X[t])) + 
    scale_color_solarized()

@@

Under appropriate choice of parameters, the model is bistable, with a low, "endemic" stable state, and a much higher stable "outbreak" state.  In the context of our beetle outbreak question, we will imagine that this model describes the population growth of the beetle over a single season, emergign from some low density following winter die off and growing subject to both it's intrinsic growth rate ($r$) and available resources ($K$) and subject to saturating density-dependent mortality imposed by tree defenses. This saturating response and the resulting threshhold dynamics in the response of affected trees exhibited by this model is quite well motivated   [e.g.@Raffa2008]: affected trees respond to beetle attack by increasing production of resins which can pitch out beetles, inhibit attraction of flying beetles to the tree, and allow other inducible defenses to activate. Initially this results in the strengthening (concave up) portion of the S curve as other defences are induced by the attack.  However, as beetle density increases further, the tree begins to exhaust resources required to produce resin and its other defences, leading to a saturating response.  Low beetle densities are thus stable, kept in check by the increasingly strong response of healthy trees.  Meanwhile, if a tree is overwhelmed, it becomes habitat for spawning many more beetles, which saturate the defences of more trees, leading to the population explosion until beetle numbers reach a much larger stable state by nearly exhausting avaliable tree habitat $K$ [@Harvey2014]. Thus our simple dynamical model from \eqref{model} captures the qualitative processes and dynamics expected in the beetle outbreak system. Bear in mind that this model is not intended to be a precise, quantitive match that could be fit directly to data, but rather, a theoretical tool to better explore the consequences of this behavior -- a realistic reflection of the process, but also simple enough to generalize to other systems dominated by similar nonlinear dynamics which can drive these threshold responses.

Models with this structure are also capable of producing long transient dynamics instead of alternative stable state dynamics by slightly changing the chosen parameter values.  Observe that by decreasing $a$, the lower stable point vanishes, but the rate of change of the system (e.g. the vertical axis, Figure 1) is still small in magnitude, and thus dynamics near this region move slowly.  This is a simple example of what @Hastings2018 has dubbed a "Ghost Attractor" -- the attractor vanishes at the fold bifurcation, but it's effects can still be felt and observed.  This could result from a weakened ability of the trees to respond to attack, which could result from additional stress such as prolonged drought or fire, both potentially intensified under changing climate.

This same model structure can thus produce rather different dynamics due to only a small change in a single parameter, as illustrated in the ensemble simulations shown in Figure 2. Each ensemble shows 50 replicate trajectories starting with the same initial condition, under parameters corresponding either to the bistable system (panel A) or the ghost attractor (panel B). Because the model is stochastic, trajectories under the bistable system have some probability of escaping the lower attractor after a sufficient interval.  Thus while it is quite easy to distinguish between the alternate stable state dynamics and the ghost attractor dynamics visually given the ensemble of replicates, given only a single replicate it could be quite difficult to be sure if it came from a particularly long transient or a stochastically escaping stable system.

@@

```{r}
P_ghost <- transition_matrix(states, actions, may(27.2), p$sigma)
P_bistable <- transition_matrix(states, actions, may(28), p$sigma)
```

@@

```{r}
x0 <- which.min(abs(states - p$x0))
Tmax <- 200
set.seed(12345)
reps <- 50
ghost_sim <- map_dfr(1:reps, function(i)
  sim(P_ghost, x0, Tmax)  %>% mutate(state = states[state], scenario = "ghost"),
  .id = "rep")
bistable_sim <- map_dfr(1:reps, function(i)
  sim(P_bistable, x0, Tmax)  %>% mutate(state = states[state], scenario = "bistable"),
  .id = "rep")

fig1_sims <- bind_rows(ghost_sim, bistable_sim)
write_csv(fig1_sims, "manuscript_cache/fig1_sims.csv")
```

```{r fig_sims}
fig1_sims %>%
  ggplot(aes(time, state, group = interaction(rep, scenario), col= scenario)) + 
  geom_line(alpha=0.3) + scale_color_solarized() + facet_wrap(~scenario)
```

This ambiguity is accentuated considerably if we are tasked with forecasting the future behavior having only observed the system during the early part of a trajectory (i.e. in the vicinity of the ghost attractor or lower stable point).  Long transients under the influence of the ghost attractor can easily be mistaken for stable node dynamics, but will have considerably greater probability of suddenly transitioning to the outbreak state than trajectories under alternate stable state dynamics in the short term.

While it is clear this situation -- in which a ghost attractor could be mistaken for a stable state -- can have important consequences for management, there has been little detailed investigation into precisely what those consequences are or how this uncertainty could be resolved. In the context of our bark beetle model, this corresponds to uncertainty as to whether the forest is healthy enough to contain the currently observed level of beetle activity to an endemic state, or whether the forest has been stressed sufficiently to be headed for a full-scale outbreak and thus warrant a more aggressive pest control response.

In the remaining sections, I explore two different approaches to addressing the problem of management under a potential ghost attractor.  The first considers the decision-theoretic optimal policy given uncertainty

@@

$$\max_{A_t \in \mathbf{A}} \sum_t \delta^t U(X_t, A_t)$$
where the evolution of $X_t$ is determined by Eq \eqref{model}. Because \eqref{model} is Markovian in $X_t$ (depending only on $X_t$ and no lagged terms),  this sequential decision problem is known as a Markov Decision Process, or MDP.  By discritizing \eqref{model} to a finite grid of possible states, we can represent the model as a transition matrix between possible states (population sizes) and solve the problem using Stochastic Dynamic Programming (SDP) [@Mangel1985, @Marescot2013].

@@

## Optimal planning under transient dynamics vs stable state dynamics

@@

```{r mdp_solns}
soln_ghost <- mdplearning::mdp_compute_policy(list(P_ghost), reward, discount, max_iter = 1e4, epsilon = 1e-2)
soln_bistable <-mdplearning::mdp_compute_policy(list(P_bistable), reward, discount, max_iter = 1e4, epsilon = 1e-4)

## Identical calculation using MDPtoolbox implementation
#soln_ghost <-MDPtoolbox::mdp_value_iteration(P_ghost, reward, discount, max_iter = 5e3, epsilon = 1e-5)

@@

```{r}
## With more noise, these are more similar.  With smaller noise, the ghost policy remains aggressive over a larger range
policy_plot <- 
  tibble(state = states,
         ghost = actions[soln_ghost$policy],
         bistable =  actions[soln_bistable$policy]
         ) %>%
  pivot_longer(c(ghost, bistable), 
               names_to="model",
               values_to = "action")

write_csv(policy_plot, "manuscript_cache/policy_plot.csv")
```

```{r fig_policy}

policy_plot %>%
  ggplot(aes(state, action, color = model)) + 
  geom_point(alpha=0.6) + 
  scale_color_solarized()
```

```{r}
### Consider other forecast models -- i.e. other transition functions

## How does the identity matrix perform as a decision tool?  
## Or, identity in f, but responsive in a?
```

@@

```{r mdp_solns_mixed_prior}
fifty_fifty <- mdplearning::mdp_compute_policy(list(P_ghost, P_bistable), reward, discount, 
                                             model_prior = c(0.5, 0.5),
                                             max_iter = 1e4, epsilon = 1e-2)

@@

```{r planning_sims}

@@

sims_by_prior <- 
 bind_rows(ghost = sim_ghost, 
           bistable = sim_bistable, 
           fifty_fifty = sim_fiftyfifty, 
           seventy_thirty = sim_seventythirty, 
           .id = "prior")

sims_by_prior %>% write_csv("manuscript_cache/sims_by_prior.csv")

@@

```{r fig_sims_by_prior}
sims_by_prior <- read_csv("manuscript_cache/sims_by_prior.csv")

sims_by_prior %>% 
  mutate(state = states[state], action = actions[action]) %>% 
  select(-obs,-value) %>%
  pivot_longer(c(-time, -prior, -reps)) %>%
  ggplot(aes(time, value, group = interaction(name, reps), col=name)) + 
  geom_path(alpha=0.5) + facet_wrap(~prior)
```

@@

```{r mdp_learning}
set.seed(12345)
Tmax <- 50
learning <- mdp_learning(list(P_ghost, P_bistable), reward, discount, 
                   model_prior = c(0.5, 0.5),
                   x0 = x0, Tmax = Tmax,
                  true_transition = P_ghost,
                  max_iter = 1e4, epsilon = 1e-2)

write_csv(learning$df, "manuscript_cache/learning_df.csv")
write_csv(learning$posterior, "manuscript_cache/learning_posterior.csv")
```

```{r fig_learning_sim}
 learning$df %>% 
  select(-value) %>% 
  gather(series, state, -time) %>% 
  ggplot(aes(time, states[state], color = series)) + geom_line()
```

```{r fig_learning_posterior}
model_names <- c("Ghost", "Bistable")
learning$posterior %>% 
  data.frame(time = 1:Tmax) %>%
  filter(time %in% floor(seq(1,10, length.out=5))) %>%
  gather(param, probability, -time, factor_key =TRUE) %>% 
  mutate(param = model_names[as.integer(param)]) %>% 
  ggplot(aes(x = param, y = probability, group = time, alpha = time)) + 
  geom_bar(stat="identity", position = "identity", show.legend = FALSE, fill=pal[1])
```

@@

Stochastic dynamic programming solutions were computed in R [@R] using the value iteration stochastic dynamic programming algorithm [@Mangel1985; @Marescot2013].  Generic R fucntions for the stochastic dynamic programming, including both planning over model uncertainty and adapative management for actively learning over model uncertainty, are supplied in the supplementary R package `mdplearning`, avaialable at <https://github.com/boettiger-lab/mdplearning>, v0.1.0 [@mdplearning].  Code to reproduce the analysis, simulations and figures shown here can be found at <https://github.com/cboettig/decisions-vs-transients>.
